Carrying out Latent Profile Analysis in Accordance With Best Practices: A Tutorial Using the tidyLPA R package

### 2020-01-13 paper.tex

---
title             : "Carrying out Latent Profile Analysis in Accordance With Best Practices: A Tutorial Using the tidyLPA R package"
shorttitle        : "tidyLPA"

author: 
  - name          : "Joshua Rosenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "1122 Volunteer Blvd., Knoxville, TN, 37996"
    email         : "jmrosenberg@utk.edu"
  - name          : "Caspar van Lissa"
    affiliation   : "2"
  - name          : "Jennifer Schmidt"
    affiliation   : "3"
  - name          : "Patrick Beymer"
    affiliation   : "3"
  - name          : "Daniel Anderson"
    affiliation   : "4"
  - name          : "Matthew Schell"
    affiliation   : "3"

affiliation:
  - id            : "1"
    institution   : "University of Tennessee, Knoxville"
  - id            : "2"
    institution   : "Utrecht University"
  - id            : "3"
    institution   : "Michigan State University"
  - id            : "4"
    institution   : "University of Oregon"
  - id            : "5"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

@@

abstract: |

keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : ""

bibliography      : ["r-references.bib", "citations.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Introduction

@@

> Grades are normally distributed, with $\mu$ = 75, $\sigma$ = 5.

Students, may (reasonably) ask, though, are grades *really* normally distributed?

```{r, include = FALSE}
library(simstudy)

def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")

set.seed(2716)
dx <- genData(1000, def)
```

@@

```{r, echo - FALSE}
library(tidyverse)
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, echo = FALSE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

This kind of distribution (that is *bimodal*) is not exclusive to grades. For
teachers, psychologists, researchers, and friends and family members, people are
highly-complex and not easily able to be characterized by *one* characteristic
or personality trait--and its distribution. In the social sciences, broadly, and
in psychology, particularly, a statistical method that can be used to describe
how people, in their individual particularities, may have similarities on the
basis of some set of measures through which they can be grouped in meaningful,
distinctive ways. This approach, which has a provenance in developmental (or
*person-oriented*) approaches (Bergman & El-Khouri, 1997; Magnusson &
Cairns, 1996; see Linnenbrink-Garcia & Wormington, 2017, for a recent review) is
an example of a general mixture model (Harring & Hodis, 2016; Pastor, Barron,
Miller, & Davis, 2007).

In this tutorial, we aim to describe one of the most commonly-used--and relevant
to psychologists--application of the general mixture model, to cases for which
all of the variables for which (relatively) homogeneous groups are identified
from among a (relatively) heterogeneous sample are continuos, *latent profile
analysis* (LPA). We note that scholars using the approach, especially in
developmental and educational psychology, use cluster analytic methods. Such an
approach has some similarities with LPA (CITE paper showing how k-means is
similar to the simplest model specification of LPA?), though LPA also has a
number of benefits (Linnenbrink-Garcia & Wormington, 2017), and we we focus on
LPA in this paper.

@@

# Latent Profile Analysis

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single dataset. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters - in terms of variances
and covariances and how they are the same or different across profiles - that
best characterize the different distributions. Then, this approach seeks to
assign to each observation a probability that the observation is a sample from
the population associated with each profile (or mixture component).

@@

We wish to note that challenge facing the analyst using LPA is that these
parameters and the distinct model parameterizations that can be estimated is the
different terminology used. As one example, Scrucca et al. (2017) refer to these
parameterizations not in terms of whether and how parameters are estimated, but
rather in terms of the geometric properties of the distributions that result
from particular parameterizations. Muthen and Muthen (1997-2017) and others
(Pastor et al., 2007) commonly refer to local independence to mean that the
covariances are fixed to zero (also described as the specification of the
covariance matrix as “diagonal,” because only the diagonal components, or the
variances, are estimated).

@@

<!-- I think we move this to the discussion in light of best practices -->

@@

# Specific Software Tools

## Other Tools

SPSS is a common tool to carry out cluster analyses (but not LPA). While
somewhat straightforward to carry out, particularly in SPSS’s graphical user
interface (GUI), there are some challenges to use of this approach. The GUI in
SPSS can be challenging, even for the most able analyst, to be able to document
every step with syntax, and so reproducing the entire analysis efficiently can
be a challenge, both for the analyst exploring various solutions and for the
reviewer looking to replicate this work. Additionally, SPSS is commercial
software (and is expensive), and so analysts access to it cannot carry out this
analysis.

Another common tool is MPlus (Muthen & Muthen, 2019). MPlus is a commercial tool
that provides functionality for many latent variable (and multi-level) models.
We will speak more to MPlus in the section on tidyLPA, as our software provides
an interface to both it and an open-source tool.

In R, a number of tools *can* be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
addition to following the same general approach, using tools that are designed
for Gaussian mixture modeling have other benefits, some efficiency-related (see
RMixMod, which uses compiled C++ code) and others in terms of ease-of-use (i.e.,
the plot methods built-in to RMixMod, mclust, and other tools). However, they
also have some drawbacks, in that it can be difficult to translate between the
model specifications, which are often described in terms of the geometric
properties of the multivariate distributions being estimated (i.e., “spherical,
equal volume”), rather than in terms of whether and how the means, variances,
and covariances are estimated. They also may use different default settings
(than those encountered in MPlus) in terms of the expectation-maximization
algorithm, which can make comparing results across tools challenging.

## tidyLPA

Because of the limitations in other tools, we set out to develop a tool that a)
provided sensible defaults and were easy to use, but provided the option to
access and modify all of the inputs to the model (i.e., low barrier, high
ceiling), b) interfaced to existing tools, and are able to translate between
what existing tools are capable of and what researchers and analysts
carrying-out person-oriented analyses would like to specify, c) made it easy to
carry-out fully-reproducible analyses and d) were well-documented. To do so, we 
created tidyLPA [@R-tidyLPA].

@@

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model
specifications with descriptions of what is estimated in terms of the variances
and covariances, the common names for the models (i.e., class-varying
unrestricted), and the covariance matrix associated with the parameterizationfor
the six models that are possible to be estimated on the website for tidyLPA (see
here).

## Installation

@@

``` {r, eval = F}
install.packages("tidyLPA")
```

@@

``` {r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("data-edu/tidyLPA")
```

``` {r, message = F}
library(tidyLPA)
library(dplyr)
```

## Examples

<!-- This is where we need to use an example from our own data -->

### Mclust via tidyLPA

Here is a brief example using the built-in `pisaUSA15` data set and variables
for broad interest, enjoyment, and self-efficacy. Note that we first type the
name of the data frame, followed by the unquoted names of the variables used to
create the profiles. We also specify the number of profiles and the model. See
`?estimate_profiles` for more details.

In these examples, we pass the results of one function to the next by *piping*
(using the `%>%` operator, loaded from the `dplyr` package). We pass the data to
a function that selects relevant variables, and then to `estimate_profiles`:

``` {r}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3)
```

### Mplus via tidyLPA

We can use Mplus simply by changing the package argument for
`estimate_profiles()`:

``` {r, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3, package = "MplusAutomation")
```

A simple summary of the analysis is printed to the console. 
The resulting object can be further passed down a pipeline to
other functions, such as `compare_solutions`, `get_data`, `get_fit`,
etc. This is the "tidy" part, in that the function can be embedded in a tidy
analysis pipeline.

If you have Mplus installed, you can call the version of this function that uses
MPlus in the same way, by adding the argument `package = "MplusAutomation`.

### Plotting the profiles

We can plot the profiles by piping the output to `plot_profiles()`.

``` {r}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    scale() %>%
    estimate_profiles(3) %>% 
    plot_profiles()
```

## Model specification

In addition to the number of profiles (specified with the `n_profiles`
argument), the model can be specified in terms of whether and how the variable
variances and covariances are estimated.

The models are specified by passing arguments to the `variance` and `covariance`
arguments. The possible values for these arguments are:

  - `variances`: "equal" and "zero"
  - `covariances`: "varying", "equal", and "zero"

@@

  - Equal variances and covariances fixed to 0 (Model 1)
  - Varying variances and covariances fixed to 0 (Model 2)
  - Equal variances and equal covariances (Model 3)
  - Varying variances and varying covariances (Model 6)

Two additional models (Models 4 and 5) can be fit using MPlus. More information
on the models can be found in the
[vignette](https://data-edu.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html).

@@

``` {r, eval = F}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3, 
                      variances = "varying",
                      covariances = "varying")
```

## Comparing a wide range of solutions

The function `compare_solutions()` compares the fit of several estimated models,
with varying numbers of profiles and model specifications:

``` {r, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(1:3, 
                      variances = c("equal", "varying"),
                      covariances = c("zero", "varying")) %>%
    compare_solutions(statistics = c("AIC", "BIC"))
```

## Accessing values in the output

@@

`get_data()` returns the data:

``` {r}
m <- pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3:4)
```

```{r}
get_data(m)
```

We note that `get_data()` returns data in wide format when applied to an object
of class tidyProfile (one element of a tidyLPA object), or when applied to a
tidyLPA object of length one. `get_data()` returns long format when applied to a
tidyLPA object containing multiple tidyProfile analyses (because then the wide
format does not make sense).

To transform data in the wide format into the long format, the `gather()`
function from the **tidyr** package can be used, e.g.:

``` {r}
get_data(m) %>% 
    tidyr::gather(Class_prob, Probability, contains("CPROB"))
```

# Recommendations Regarding Analytic Choices

## GRoLTs checklist

```{r, echo = FALSE}
t <- read_csv("GRoLTS-applicability-to-LPA.csv")
tp <- t %>% 
  filter(use == 1) %>% 
  mutate(`LPA Number` = 1:nrow(.)) %>% 
  select(`GRoLTS Number` = Number, `LPA Number`, `Checklist Item`)

tp %>% 
  DT::datatable()
```

@@

### Is the missing data mechanism reported?

### Is a description provided of what variables are related to attrition/missing data?

### Is a description provided of how missing data in the analyses were dealt with?

### Is information about the distribution of the observed variables included?

### Is the software mentioned?

### Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented?

### Is information reported about the number of random start values and final iterations included?

### Are the model comparison (and selection) tools described from a statistical perspective?

### Are the total number of fitted models reported, including a one-class solution?

### Are the number of cases per class reported for each model (absolute sample size, or proportion)?

### If classification of cases in a trajectory is the goal, is entropy reported?

### Are all of the estimates plotted?

### Are the raw data plotted along with the estimates?

### Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?

### Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?

There are a number of analytic choices that need to be made when carrying out
person-oriented analyses. Because such person-oriented approaches are often more
subjective (in practice) than other approaches (Linnenbrink-Garcia and
Wormington, 2017), there is no one rule for determining the solution obtained.
This solution is obtained on the basis of multiple decisions, such as the number
of profiles selected or the modeling decisions such as what specific options are
used for the cluster analysis (i.e., the distance metric used to calculate the
similarity of the observations as part of the Ward’s hierarchical clustering) or
what parameters are estimated and how as part of LPA.

@@

## How the Data are Centering or Scaled

@@

## How to Choose the Number of Profiles/Classes

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Carrying out Latent Profile Analysis in Accordance With Best Practices: A Tutorial Using the tidyLPA R package},
            pdfauthor={Joshua Rosenberg, Caspar van Lissa, Jennifer Schmidt, Patrick Beymer, Daniel Anderson, \& Matthew Schell},
            pdfkeywords={Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother

\title{Carrying out Latent Profile Analysis in Accordance With Best Practices: A Tutorial Using the tidyLPA R package}
\author{Joshua Rosenberg\textsuperscript{1}, Caspar van Lissa\textsuperscript{2}, Jennifer Schmidt\textsuperscript{3}, Patrick Beymer\textsuperscript{3}, Daniel Anderson\textsuperscript{4}, \& Matthew Schell\textsuperscript{3}}
\date{}

\authornote{Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

@@

Correspondence concerning this article should be addressed to Joshua Rosenberg, 1122 Volunteer Blvd., Knoxville, TN, 37996. E-mail: \href{mailto:jmrosenberg@utk.edu}{\nolinkurl{jmrosenberg@utk.edu}}}

\abstract{

}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\begin{quote}
Grades are normally distributed, with \(\mu\) = 75, \(\sigma\) = 5.
\end{quote}

Students, may (reasonably) ask, though, are grades \emph{really} normally distributed?

@@

\includegraphics{paper_files/figure-latex/echo - FALSE-1.pdf}

@@

\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}

This kind of distribution (that is \emph{bimodal}) is not exclusive to grades. For
teachers, psychologists, researchers, and friends and family members, people are
highly-complex and not easily able to be characterized by \emph{one} characteristic
or personality trait--and its distribution. In the social sciences, broadly, and
in psychology, particularly, a statistical method that can be used to describe
how people, in their individual particularities, may have similarities on the
basis of some set of measures through which they can be grouped in meaningful,
distinctive ways. This approach, which has a provenance in developmental (or
\emph{person-oriented}) approaches (Bergman \& El-Khouri, 1997; Magnusson \&
Cairns, 1996; see Linnenbrink-Garcia \& Wormington, 2017, for a recent review) is
an example of a general mixture model (Harring \& Hodis, 2016; Pastor, Barron,
Miller, \& Davis, 2007).

In this tutorial, we aim to describe one of the most commonly-used--and relevant
to psychologists--application of the general mixture model, to cases for which
all of the variables for which (relatively) homogeneous groups are identified
from among a (relatively) heterogeneous sample are continuos, \emph{latent profile
analysis} (LPA). We note that scholars using the approach, especially in
developmental and educational psychology, use cluster analytic methods. Such an
approach has some similarities with LPA (CITE paper showing how k-means is
similar to the simplest model specification of LPA?), though LPA also has a
number of benefits (Linnenbrink-Garcia \& Wormington, 2017), and we we focus on
LPA in this paper.

@@

\hypertarget{latent-profile-analysis}{%
\section{Latent Profile Analysis}\label{latent-profile-analysis}}

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single dataset. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters - in terms of variances
and covariances and how they are the same or different across profiles - that
best characterize the different distributions. Then, this approach seeks to
assign to each observation a probability that the observation is a sample from
the population associated with each profile (or mixture component).

@@

We wish to note that challenge facing the analyst using LPA is that these
parameters and the distinct model parameterizations that can be estimated is the
different terminology used. As one example, Scrucca et al. (2017) refer to these
parameterizations not in terms of whether and how parameters are estimated, but
rather in terms of the geometric properties of the distributions that result
from particular parameterizations. Muthen and Muthen (1997-2017) and others
(Pastor et al., 2007) commonly refer to local independence to mean that the
covariances are fixed to zero (also described as the specification of the
covariance matrix as ``diagonal,'' because only the diagonal components, or the
variances, are estimated).

@@

@@

\hypertarget{specific-software-tools}{%
\section{Specific Software Tools}\label{specific-software-tools}}

\hypertarget{other-tools}{%
\subsection{Other Tools}\label{other-tools}}

SPSS is a common tool to carry out cluster analyses (but not LPA). While
somewhat straightforward to carry out, particularly in SPSS's graphical user
interface (GUI), there are some challenges to use of this approach. The GUI in
SPSS can be challenging, even for the most able analyst, to be able to document
every step with syntax, and so reproducing the entire analysis efficiently can
be a challenge, both for the analyst exploring various solutions and for the
reviewer looking to replicate this work. Additionally, SPSS is commercial
software (and is expensive), and so analysts access to it cannot carry out this
analysis.

Another common tool is MPlus (Muthen \& Muthen, 2019). MPlus is a commercial tool
that provides functionality for many latent variable (and multi-level) models.
We will speak more to MPlus in the section on tidyLPA, as our software provides
an interface to both it and an open-source tool.

In R, a number of tools \emph{can} be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
addition to following the same general approach, using tools that are designed
for Gaussian mixture modeling have other benefits, some efficiency-related (see
RMixMod, which uses compiled C++ code) and others in terms of ease-of-use (i.e.,
the plot methods built-in to RMixMod, mclust, and other tools). However, they
also have some drawbacks, in that it can be difficult to translate between the
model specifications, which are often described in terms of the geometric
properties of the multivariate distributions being estimated (i.e., ``spherical,
equal volume''), rather than in terms of whether and how the means, variances,
and covariances are estimated. They also may use different default settings
(than those encountered in MPlus) in terms of the expectation-maximization
algorithm, which can make comparing results across tools challenging.

\hypertarget{tidylpa}{%
\subsection{tidyLPA}\label{tidylpa}}

Because of the limitations in other tools, we set out to develop a tool that a)
provided sensible defaults and were easy to use, but provided the option to
access and modify all of the inputs to the model (i.e., low barrier, high
ceiling), b) interfaced to existing tools, and are able to translate between
what existing tools are capable of and what researchers and analysts
carrying-out person-oriented analyses would like to specify, c) made it easy to
carry-out fully-reproducible analyses and d) were well-documented. To do so, we
created tidyLPA (Rosenberg, Beymer, Anderson, Lissa, \& Schmidt, 2018).

@@

One way that tidyLPA is designed to be easy to use is that it assumes a ``tidy''
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen \& Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model
specifications with descriptions of what is estimated in terms of the variances
and covariances, the common names for the models (i.e., class-varying
unrestricted), and the covariance matrix associated with the parameterizationfor
the six models that are possible to be estimated on the website for tidyLPA (see
here).

\hypertarget{installation}{%
\subsection{Installation}\label{installation}}

@@

@@

\hypertarget{examples}{%
\subsection{Examples}\label{examples}}

\hypertarget{mclust-via-tidylpa}{%
\subsubsection{Mclust via tidyLPA}\label{mclust-via-tidylpa}}

Here is a brief example using the built-in \texttt{pisaUSA15} data set and variables
for broad interest, enjoyment, and self-efficacy. Note that we first type the
name of the data frame, followed by the unquoted names of the variables used to
create the profiles. We also specify the number of profiles and the model. See
\texttt{?estimate\_profiles} for more details.

In these examples, we pass the results of one function to the next by \emph{piping}
(using the \texttt{\%\textgreater{}\%} operator, loaded from the \texttt{dplyr} package). We pass the data to
a function that selects relevant variables, and then to \texttt{estimate\_profiles}:

\begin{verbatim}
## tidyLPA analysis using mclust: 
## 
##  Model Classes AIC    BIC    Entropy prob_min prob_max n_min n_max BLRT_p
##  1     3       629.10 665.57 0.82    0.83     0.96     0.03  0.70  0.01
\end{verbatim}

\hypertarget{mplus-via-tidylpa}{%
\subsubsection{Mplus via tidyLPA}\label{mplus-via-tidylpa}}

We can use Mplus simply by changing the package argument for
\texttt{estimate\_profiles()}:

A simple summary of the analysis is printed to the console.
The resulting object can be further passed down a pipeline to
other functions, such as \texttt{compare\_solutions}, \texttt{get\_data}, \texttt{get\_fit},
etc. This is the ``tidy'' part, in that the function can be embedded in a tidy
analysis pipeline.

If you have Mplus installed, you can call the version of this function that uses
MPlus in the same way, by adding the argument \texttt{package\ =\ "MplusAutomation}.

\hypertarget{plotting-the-profiles}{%
\subsubsection{Plotting the profiles}\label{plotting-the-profiles}}

We can plot the profiles by piping the output to \texttt{plot\_profiles()}.

\includegraphics{paper_files/figure-latex/unnamed-chunk-7-1.pdf}

\hypertarget{model-specification}{%
\subsection{Model specification}\label{model-specification}}

In addition to the number of profiles (specified with the \texttt{n\_profiles}
argument), the model can be specified in terms of whether and how the variable
variances and covariances are estimated.

The models are specified by passing arguments to the \texttt{variance} and \texttt{covariance}
arguments. The possible values for these arguments are:

\begin{itemize}
\tightlist
\item
  \texttt{variances}: ``equal'' and ``zero''
\item
  \texttt{covariances}: ``varying'', ``equal'', and ``zero''
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Equal variances and covariances fixed to 0 (Model 1)
\item
  Varying variances and covariances fixed to 0 (Model 2)
\item
  Equal variances and equal covariances (Model 3)
\item
  Varying variances and varying covariances (Model 6)
\end{itemize}

Two additional models (Models 4 and 5) can be fit using MPlus. More information
on the models can be found in the
\href{https://data-edu.github.io/tidyLPA/articles/Introduction_to_tidyLPA.html}{vignette}.

@@

\hypertarget{comparing-a-wide-range-of-solutions}{%
\subsection{Comparing a wide range of solutions}\label{comparing-a-wide-range-of-solutions}}

The function \texttt{compare\_solutions()} compares the fit of several estimated models,
with varying numbers of profiles and model specifications:

\hypertarget{accessing-values-in-the-output}{%
\subsection{Accessing values in the output}\label{accessing-values-in-the-output}}

@@

\texttt{get\_data()} returns the data:

\begin{verbatim}
## # A tibble: 700 x 9
##    model_number classes_number broad_interest enjoyment self_efficacy Class
##           <dbl>          <int>          <dbl>     <dbl>         <dbl> <dbl>
##  1            1              3            3.8       4            1        1
##  2            1              3            3         3            2.75     3
##  3            1              3            1.8       2.8          3.38     3
##  4            1              3            1.4       1            2.75     2
##  5            1              3            1.8       2.2          2        3
##  6            1              3            1.6       1.6          1.88     3
##  7            1              3            3         3.8          2.25     1
##  8            1              3            2.6       2.2          2        3
##  9            1              3            1         2.8          2.62     3
## 10            1              3            2.2       2            1.75     3
## # … with 690 more rows, and 3 more variables: Class_prob <int>,
## #   Probability <dbl>, id <int>
\end{verbatim}

We note that \texttt{get\_data()} returns data in wide format when applied to an object
of class tidyProfile (one element of a tidyLPA object), or when applied to a
tidyLPA object of length one. \texttt{get\_data()} returns long format when applied to a
tidyLPA object containing multiple tidyProfile analyses (because then the wide
format does not make sense).

To transform data in the wide format into the long format, the \texttt{gather()}
function from the \textbf{tidyr} package can be used, e.g.:

\begin{verbatim}
## # A tibble: 700 x 9
##    model_number classes_number broad_interest enjoyment self_efficacy Class
##           <dbl>          <int>          <dbl>     <dbl>         <dbl> <dbl>
##  1            1              3            3.8       4            1        1
##  2            1              3            3         3            2.75     3
##  3            1              3            1.8       2.8          3.38     3
##  4            1              3            1.4       1            2.75     2
##  5            1              3            1.8       2.2          2        3
##  6            1              3            1.6       1.6          1.88     3
##  7            1              3            3         3.8          2.25     1
##  8            1              3            2.6       2.2          2        3
##  9            1              3            1         2.8          2.62     3
## 10            1              3            2.2       2            1.75     3
## # … with 690 more rows, and 3 more variables: Class_prob <int>,
## #   Probability <dbl>, id <int>
\end{verbatim}

\hypertarget{recommendations-regarding-analytic-choices}{%
\section{Recommendations Regarding Analytic Choices}\label{recommendations-regarding-analytic-choices}}

\hypertarget{grolts-checklist}{%
\subsection{GRoLTs checklist}\label{grolts-checklist}}

\includegraphics{paper_files/figure-latex/unnamed-chunk-13-1.pdf}

@@

\hypertarget{is-the-missing-data-mechanism-reported}{%
\subsubsection{Is the missing data mechanism reported?}\label{is-the-missing-data-mechanism-reported}}

\hypertarget{is-a-description-provided-of-what-variables-are-related-to-attritionmissing-data}{%
\subsubsection{Is a description provided of what variables are related to attrition/missing data?}\label{is-a-description-provided-of-what-variables-are-related-to-attritionmissing-data}}

\hypertarget{is-a-description-provided-of-how-missing-data-in-the-analyses-were-dealt-with}{%
\subsubsection{Is a description provided of how missing data in the analyses were dealt with?}\label{is-a-description-provided-of-how-missing-data-in-the-analyses-were-dealt-with}}

\hypertarget{is-information-about-the-distribution-of-the-observed-variables-included}{%
\subsubsection{Is information about the distribution of the observed variables included?}\label{is-information-about-the-distribution-of-the-observed-variables-included}}

\hypertarget{is-the-software-mentioned}{%
\subsubsection{Is the software mentioned?}\label{is-the-software-mentioned}}

\hypertarget{are-alternative-specifications-of-the-between-class-differences-in-variancecovariance-matrix-structure-considered-and-clearly-documented}{%
\subsubsection{Are alternative specifications of the between-class differences in variance--covariance matrix structure considered and clearly documented?}\label{are-alternative-specifications-of-the-between-class-differences-in-variancecovariance-matrix-structure-considered-and-clearly-documented}}

\hypertarget{is-information-reported-about-the-number-of-random-start-values-and-final-iterations-included}{%
\subsubsection{Is information reported about the number of random start values and final iterations included?}\label{is-information-reported-about-the-number-of-random-start-values-and-final-iterations-included}}

\hypertarget{are-the-model-comparison-and-selection-tools-described-from-a-statistical-perspective}{%
\subsubsection{Are the model comparison (and selection) tools described from a statistical perspective?}\label{are-the-model-comparison-and-selection-tools-described-from-a-statistical-perspective}}

\hypertarget{are-the-total-number-of-fitted-models-reported-including-a-one-class-solution}{%
\subsubsection{Are the total number of fitted models reported, including a one-class solution?}\label{are-the-total-number-of-fitted-models-reported-including-a-one-class-solution}}

\hypertarget{are-the-number-of-cases-per-class-reported-for-each-model-absolute-sample-size-or-proportion}{%
\subsubsection{Are the number of cases per class reported for each model (absolute sample size, or proportion)?}\label{are-the-number-of-cases-per-class-reported-for-each-model-absolute-sample-size-or-proportion}}

\hypertarget{if-classification-of-cases-in-a-trajectory-is-the-goal-is-entropy-reported}{%
\subsubsection{If classification of cases in a trajectory is the goal, is entropy reported?}\label{if-classification-of-cases-in-a-trajectory-is-the-goal-is-entropy-reported}}

\hypertarget{are-all-of-the-estimates-plotted}{%
\subsubsection{Are all of the estimates plotted?}\label{are-all-of-the-estimates-plotted}}

\hypertarget{are-the-raw-data-plotted-along-with-the-estimates}{%
\subsubsection{Are the raw data plotted along with the estimates?}\label{are-the-raw-data-plotted-along-with-the-estimates}}

\hypertarget{are-characteristics-of-the-final-class-solution-numerically-described-i.e.-means-sdse-n-ci-etc.}{%
\subsubsection{Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?}\label{are-characteristics-of-the-final-class-solution-numerically-described-i.e.-means-sdse-n-ci-etc.}}

\hypertarget{are-the-syntax-files-available-either-in-the-appendix-supplementary-materials-or-from-the-authors}{%
\subsubsection{Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?}\label{are-the-syntax-files-available-either-in-the-appendix-supplementary-materials-or-from-the-authors}}

There are a number of analytic choices that need to be made when carrying out
person-oriented analyses. Because such person-oriented approaches are often more
subjective (in practice) than other approaches (Linnenbrink-Garcia and
Wormington, 2017), there is no one rule for determining the solution obtained.
This solution is obtained on the basis of multiple decisions, such as the number
of profiles selected or the modeling decisions such as what specific options are
used for the cluster analysis (i.e., the distance metric used to calculate the
similarity of the observations as part of the Ward's hierarchical clustering) or
what parameters are estimated and how as part of LPA.

@@

\hypertarget{how-the-data-are-centering-or-scaled}{%
\subsection{How the Data are Centering or Scaled}\label{how-the-data-are-centering-or-scaled}}

@@

\hypertarget{how-to-choose-the-number-of-profilesclasses}{%
\subsection{How to Choose the Number of Profiles/Classes}\label{how-to-choose-the-number-of-profilesclasses}}
